
Load the relevant packages.
```{r, echo = TRUE, message= FALSE, warning = FALSE}
library("jrModelling")
library("broom")
library("tidyverse")
```

\noindent I've starred, $^*$, some of the questions. This indicates that we didn't directly cover the material in the lecture. If you are particular interested in this statistical area, try the question. Otherwise, just move on.

## Question 1

We conducted an experiment and collected the data in the tables above. This data set isn't paired.

\begin{table}
  \centering
  \begin{tabular}{@{}llllll@{}}
    \toprule
    \multicolumn{6}{l}{Method A}\\
    \midrule
    78.64 & 79.01 & 79.57 & 79.52 & 80.71 & 79.95\\
    78.50 & 79.10 & 81.98 & 80.09 & 80.29 & 80.22\\
    \bottomrule
  \end{tabular}
\end{table}
\begin{table}
  \centering
  \begin{tabular}{@{}llllll@{}}
    \toprule
    \multicolumn{6}{l}{Method B}\\
    \midrule
     81.92 & 81.12 & 82.47 & 82.86 & 82.89 & 82.45 \\
     82.51 & 81.11 & 83.07 & 82.77 & 82.38 & 83.14 \\
    \bottomrule
  \end{tabular}
\end{table}

Use the code below to read in the data set to R. If you prefer, you can copy the data into a `.csv` file, and practice reading in data using `read_csv()`.

```{r, message = FALSE, warning = FALSE, echo = TRUE}
a = c(78.64,79.01, 79.57, 79.52, 80.71, 79.95, 78.50,
  79.10, 81.98, 80.09, 80.29, 80.22)
b = c(81.92, 81.12, 82.47, 82.86, 82.89, 82.45,
     82.51, 81.11, 83.07, 82.77, 82.38, 83.14)
results = tibble(a, b)
```



a) Sometimes, it's useful to get data into a tidy format before we start analysis. We can then use the `gather()` function from **tidyr** to get the data into the correct shape for analysis. The `gather()` function gathers up the two method columns into one new variable "method".  If you haven't used the `gather()` function before, have a look at the help file `?gather()`. 

    ```{r, echo = TRUE}
 results = gather(results, key = "method", value = "value")
    ```

b) Exploratory data analysis. - Construct boxplots, density plots and q-q plots for both data sets. Work out the means and standard deviations. Before carrying out any statistical test, what do you think your conclusions will be? Do you think the variances are roughly equal? Do you think the data conforms to a normal distribution.

```{r, fig.keep = 'none'}

# Box plot
ggplot(results, aes(x = method, y = value)) + 
  geom_boxplot()

# Density plot
ggplot(results, aes(x = value, fill = method)) + 
  geom_density(alpha = 0.4)

# QQ plots
ggplot(results, aes(sample = value, col = method)) + 
  geom_qq() + 
  geom_qq_line()

# Means and standard deviations
results %>%
  group_by(method) %>%
    summarise(mean = round(mean(value),2 ),
              sd = round(sd(value), 2))
```


c) Carry out a two sample $t$-test. Assume that the variances are unequal.
    ```{r,  }
    t.test(value ~ method, data = results, var.equal=FALSE)
    ```
How does this answer compare with your intuition?

d) Carry out a two sample $t$-test, assuming equal variances.
    ```{r,  }
    t.test(value ~ method, data = results, var.equal=TRUE)
    ```

e) $^*$ Now carry out a `wilcox.test()`.
```{r, }
wilcox.test(value ~ method, data = results)
```

## Question 2

\newthought{Suppose we} are interested whether successful business executives are affected by their zodiac sign. We have collected 4265 samples and obtained the data shown in table \ref{tab:1}.

\begin{table*}[h]
  \resizebox{\textwidth}{!}{%
  \centering
  \begin{tabular}{@{}llll llll llll@{}}
    \toprule
    Aries & Taurus & Gemini & Cancer & Leo & Virgo & Libra & Scorpio &
    Sagittarius & Capricorn & Aquarius & Pisces \\
    \midrule
    348 & 353 & 359 & 357 & 350 & 355 & 359 & 367 & 345 & 362 & 343 & 367\\
\bottomrule
  \end{tabular}}
  \caption{Zodiac signs of 4265 business executives}
  \label{tab:1}
\end{table*}

Load the data into R using the following code
```{r, echo = TRUE}
zodiac = c(348, 353, 359, 357, 350, 355, 359, 367, 345, 362, 343, 367)
```

a) Carry out a $\chi^2$ goodness of fit test on the zodiac data. Are business
executives distributed uniformly across zodiac signs?

    ```{r,  }
    zodiac = c(348, 353, 359, 357, 350, 355, 359, 367, 345, 362, 343, 367)
    m = chisq.test(zodiac)
    ## Since p > 0.05 we can't accept the alternative hypothesis.
    ## However, the question is worded as though we can "prove" the Null
    ## hypotheis, which we obviously can't do.
    ```


b) What are the expected values for each zodiac sign? Hint: Use `augment()`
    ```{r, message = FALSE, warning = FALSE }
    library("broom")
    m_aug = augment(m)
    m_aug$.expected
    ```

c) Use `augment()` to look at the standardised residuals. Which are large?
    ```{r,  }
    m_aug$.stdres
    ```

## Question 3

\newthought{The} University of Texas Southwestern Medical Center examined
whether the risk of contracting Hepatitis C was related to tattoo
use.^[Haley, R. and Fischer, P.R. 2001] The data from the study is
summarised as follows:

\begin{table}[h]
  \centering
  \caption{Counts of patients by their Hepatitis C status and whether they had a
    tattoo from a parlour, from elsewhere or had no tattoo at all.}
\begin{tabular}{@{}llll@{}}
  \toprule
  & Hepatitis C & No Hepatitis C & Total \\
  \midrule
  Tattoo, Parlour & 17 & 35 & 52 \\
  Tattoo, elsewhere & 8 & 53  & 61 \\
  No tattoo & 22 & 491 & 513\\
  \midrule
  Total & 47 & 579 & 626 \\
  \bottomrule
\end{tabular}
\end{table}

Use the following code to create the data set.

```{r, echo = TRUE}
tattoo = tibble(hep = c(17, 8, 22), no_hep = c(35, 53, 491))
```

a) Carry out a $\chi^2$ test to determine if the Hepatitis is related to
tattoo status.
```{r, warning=FALSE, }
    m = chisq.test(tattoo)
```

b) When carrying out $\chi^2$ tests, we should make sure that individual
cells have expected values of at least five, otherwise the distributional
assumptions may be invalid. What are the expected values of each cell. Which
cells have an expected value less than five?
    ```{r,  }
    m_aug = augment(m)
    m_aug$.expected
    ```

c) Since some of the cells have expected values slightly less than five, we
should ensure that these aren't driving the test statistic. Look at the test
residuals. Which residuals are large? What should you do now?
    ```{r,  }
    ## Look at the .stdres column of aug_m. Some residuals are very large.
    ## We could consider combining cells.
    ```

## Solutions

Solutions are contained within this package:
```{r, echo = TRUE, eval = FALSE}
vignette("solutions1", package = "jrModelling")
```
